---
title: "Introduction to CellChatCustomPlots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to CellChatCustomPlots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>"
)
library(ggplot2)
library(dplyr)
```

## Overview

`CellChatCustomPlots` provides custom heatmap and bubble plot visualizations
for CellChat objects to explore cell-cell communication patterns across
single or multiple samples.

### Functions

| Function | Description |
|----------|-------------|
| `prephm()` | Preprocess CellChat data for heatmap |
| `makehm()` | Heatmap for a single CellChat object |
| `makehm_pooled()` | Pooled heatmap across multiple objects |
| `makehm_pooled_onepathway()` | Pooled heatmap for one pathway |
| `extract_cellchat_data()` | Extract LR interaction data across objects |

---

## Load the Package and Example Data

```{r load}
library(CellChatCustomPlots)

data("example_CellChat_list")
data("example_DB")

length(example_CellChat_list)
names(example_CellChat_list)

# Available cell types
dimnames(example_CellChat_list[[1]]@netP$prob)[[1]]

# Available pathways
dimnames(example_CellChat_list[[1]]@netP$prob)[[3]]
```

## Single Sample Heatmap

```{r sshm, include = TRUE, fig.retina=2, out.width="100%", fig.width=10, fig.height=6, dev='png'}

library(circlize)

# Define sources and targets
sources <- c("mCAF1", "Basal", "CD8_Prolif", "Vendo1")  # replace with your cell types
targets <- c("mCAF1", "Basal","CD8_Prolif", "Vendo1")  # replace with your cell types

# Define color scale
col_fun = circlize::colorRamp2(c(0, 0.005, 1), c("white", "darkseagreen2", "darkgreen"))

result <- makehm(
  cellchatobj = example_CellChat_list[[1]],
  col_fun     = col_fun,
  sources     = sources,
  targets     = targets,
  fontsize    = 10,
  hmtitle     = "Sample Heatmap"
)

# Draw the heatmap
ComplexHeatmap::draw(result[[2]])

# Inspect the data matrix
head(result[[3]])
```

## Multi Sample Heatmap

```{r mshm, include = TRUE, fig.retina=2, out.width="100%", fig.width=10, fig.height=6, dev='png'}
result_pooled <- makehm_pooled(
  all_objects_list = example_CellChat_list,
  col_fun          = col_fun,
  sources          = sources,
  targets          = targets,
  fontsize         = 10,
  hmtitle          = "Pooled Heatmap"
)

ComplexHeatmap::draw(result_pooled[[2]])
```

## Multi Sample Heatmap One Pathway

```{r mshmOP, include = TRUE, eval = TRUE, fig.retina=2, out.width="100%", fig.width=10, fig.height=6, dev='png'}

# Use first available pathway
pathway_of_interest <- dimnames(example_CellChat_list[[1]]@netP$prob)[[3]][1]

result_pathway <- makehm_pooled_onepathway(
  all_objects_list = example_CellChat_list,
  col_fun          = col_fun,
  sources          = sources,
  targets          = targets,
  fontsize         = 10,
  pathway          = pathway_of_interest,
  hmtitle          = "Single Pathway Heatmap"
)

ComplexHeatmap::draw(result_pathway[[2]])
```

## Bubble Plot

```{r bubbleplot, include = TRUE, fig.retina=2, out.width="100%", fig.width=10, fig.height=6, dev='png'}

# Re-attach DB before using extract_cellchat_data
objects_with_db <- lapply(example_CellChat_list, function(obj) {
  obj@DB <- example_DB
  return(obj)
})

lr_data <- extract_cellchat_data(
  all_objects_list = objects_with_db,
  sources.use      = sources,
  targets.use      = targets,
  signaling        = "CXCL" #dimnames(example_CellChat_list[[1]]@netP$prob)[[3]][1]
)

head(lr_data)

#plot
bubble_data <- lr_data %>%
  group_by(source, target, ligand, receptor, interaction_name) %>%
  summarise(
    prob_avg = mean(prob.original, na.rm = TRUE),
    n_patients = n_distinct(patient),
    .groups = 'drop'
  ) %>%
  dplyr::filter(prob_avg > 0, n_patients > 0)

# Create interaction label
bubble_data$interaction_label <- paste0(bubble_data$ligand, "-", bubble_data$receptor)

# Bubble plot: size = # patients, color = avg prob
p1 <- ggplot(bubble_data, 
             aes(#x = reorder(interaction_label, prob_avg), 
               x = interaction_label,
                 y = paste(source, "→", target), 
                 size = n_patients,      # Size by # patients
                 fill = prob_avg)) +     # Color by avg probability
  geom_point(shape = 21, stroke = 0.8, alpha = 0.85) +
  scale_size_continuous(range = c(1, 5), name = "# Patients") +
  scale_fill_viridis_c(name = "Avg Probability", option = "plasma") +
  labs(title = "",
       subtitle = "Pooled across patients",
       x = "Ligand-Receptor Pair", y = "Source → Target") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
        axis.text.y = element_text(size = 11, face = "bold"),
        legend.position = "right") 

print(p1)

```

## Create Small Toy CellChat-Like Objects

The heatmap functions use the `@netP$prob` slot from CellChat objects. For
small examples, you can create tiny CellChat-like S4 objects with only that
slot.

```{r toy-data}
setClass("ToyCellChat", slots = c(netP = "list"))

make_toy_cellchat <- function(values,
                              sources = c("Basal", "mCAF1", "Vendo1"),
                              targets = c("Basal", "mCAF1", "Vendo1"),
                              pathways = c("CXCL", "COLLAGEN", "LAMININ", "MHC-I")) {
  prob <- array(
    0,
    dim = c(length(sources), length(targets), length(pathways)),
    dimnames = list(
      source = sources,
      target = targets,
      pathway_name = pathways
    )
  )

  for (value in values) {
    prob[value$source, value$target, value$pathway] <- value$prob
  }

  new("ToyCellChat", netP = list(prob = prob))
}

young_objects <- list(
  young1 = make_toy_cellchat(list(
    list(source = "Basal", target = "mCAF1", pathway = "CXCL", prob = 0.10),
    list(source = "Basal", target = "Vendo1", pathway = "COLLAGEN", prob = 0.05),
    list(source = "mCAF1", target = "Basal", pathway = "LAMININ", prob = 0.12)
  )),
  young2 = make_toy_cellchat(list(
    list(source = "Basal", target = "mCAF1", pathway = "CXCL", prob = 0.14),
    list(source = "Basal", target = "Vendo1", pathway = "COLLAGEN", prob = 0.06),
    list(source = "mCAF1", target = "Basal", pathway = "LAMININ", prob = 0.10)
  ))
)

old_objects <- list(
  old1 = make_toy_cellchat(list(
    list(source = "Basal", target = "mCAF1", pathway = "CXCL", prob = 0.30),
    list(source = "Basal", target = "Vendo1", pathway = "COLLAGEN", prob = 0.03),
    list(source = "mCAF1", target = "Basal", pathway = "LAMININ", prob = 0.04),
    list(source = "Vendo1", target = "mCAF1", pathway = "MHC-I", prob = 0.15)
  )),
  old2 = make_toy_cellchat(list(
    list(source = "Basal", target = "mCAF1", pathway = "CXCL", prob = 0.26),
    list(source = "Basal", target = "Vendo1", pathway = "COLLAGEN", prob = 0.02),
    list(source = "mCAF1", target = "Basal", pathway = "LAMININ", prob = 0.05),
    list(source = "Vendo1", target = "mCAF1", pathway = "MHC-I", prob = 0.12)
  ))
)

toy_sources <- c("Basal", "mCAF1")
toy_targets <- c("mCAF1", "Vendo1")
```

## Differential Heatmap Between Conditions

```{r diffhm, include = TRUE, fig.retina=2, out.width="100%", fig.width=8, fig.height=5, dev='png'}
result_diff <- makehm_compareconditions(
  list_ctrl  = young_objects,
  list_treat = old_objects,
  sources    = toy_sources,
  targets    = toy_targets,
  fontsize   = 10,
  hmtitle    = "Old vs Young",
  min_rowsum = 0.01,
  row_order  = "absolute_change"
)

ComplexHeatmap::draw(result_diff$heatmap)

result_diff$diff_mat
```

## Customizing Heatmaps

```{r customization, include = TRUE, fig.retina=2, out.width="100%", fig.width=8, fig.height=5, dev='png'}
result_custom <- makehm_compareconditions(
  list_ctrl        = young_objects,
  list_treat       = old_objects,
  sources          = toy_sources,
  targets          = toy_targets,
  fontsize         = 10,
  hmtitle          = "Customized Difference Heatmap",
  min_rowsum       = 0.01,
  neg_col          = "navy",
  mid_col          = "white",
  pos_col          = "red",
  row_order        = "treat_increased",
  show_row_barplot = TRUE,
  show_col_barplot = TRUE,
  show_row_labels  = TRUE
)

ComplexHeatmap::draw(result_custom$heatmap)
```
